On Optimality of Finite Element Integration

### 2015-11-02 TOMS_Paper/assembly.tex
Guess what? Still on fuckin section 2

\subsection{Quadrature Mode}
Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies that a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Tensor Contraction Mode}
\label{sec:tc}
Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. Because of only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated ``reference tensor'' $A_{i_1 i_2 \alpha}$ and the cell-dependent ``geometry tensor'' $G_{K}^\alpha$ are exposed.

@@

\subsection{Loop Nests and Optimality}
In order to make the article self-contained, we start with reviewing basic compiler terminology.

@@

\subsection{Sharing Elimination}
\label{sec:sharing-elimination}
We start with introducing the fundamental notion of sharing.

@@

Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Note that this normal form respects Definition~\ref{def:fem-loopnest}. We use the normal form for the following considerations:

\begin{enumerate}
\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$. That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$). By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
\item Let us now consider the general case in which test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}). Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest. This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost). Two observations follow:
\begin{enumerate}
\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
\item However, the difference between the two tends to be negligible, since the operation count within the outer product loop nest is identical.
\end{enumerate}
\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among terms. Then, factorizing any $f_i \in \sigma_{ei}$ \textit{across two terms} $w1$ \textit{and} $w2$ always increases the operation count with respect to points 1) or 2). This is because of two facts. Firstly, the number of operations at every iteration is larger. Secondly, sub-expressions that are not hoistable due to the inherent dependencies on loop indices are exposed. On the other hand, factorizing any $f_i \in \sigma_{ei}$ \textit{within a generic term} $w$ promotes hoisting of geometry-dependent sub-expressions at the level of $L_e$. 
\end{enumerate}
\item Inherent in the previous points is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for maximizing code motion. The fact that expressions result from sequences of inner products and loop nests adhere to a specific structure allows us to define a rigid transformation process leading, in general, to operation count minimization. 
\item There only are a few corner circumstances in which point 4) decades. Briefly:
\begin{enumerate}
\item As already explained, the dichotomy between points 1) and 2).
\item The case in which loops degenerate (i.e., their trip counts is 1), which makes code motion irrelevant and certain factorizations sub-optimal (e.g., the factorization $(a f_i + b g_i) (c f_i + d f_i)$ is optimal if and only if the size of $L_i$ is 1).
\end{enumerate}
\end{enumerate}

These considerations can easily be turned into a constructive proof showing that any loop nest produced by Algorithm~\ref{algo:opt} is optimal if and only if: (i) we restrict, as specified at the beginning of the section, to factorization, code motion and common sub-expression elimination (what we mean by this is also clarified in the next section); (ii) we neglect the pathological cases in point 5) above.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction. Assume then we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
\end{Algo}

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and/or common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.

@@

Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. However, we can make use of a result -- the foundation of tensor contraction mode -- to simplify our task. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors. This result can be turned into a transformation algorithm for loops and expressions, for which we provide a succinct description below.

@@

\subsection{Transformation Algorithm}

@@

\subsection{Automation through the COFFEE Language}
COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality). In other words, COFFEE only exploits domain properties, and it is not specifically tied to finite element in any way. Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.

@@

\subsection{Independence from Form Compilers}
\label{sec:zeros}
One of the key pillars of COFFEE is independence from form compilers. Being capable of handling generic abstract syntax trees, COFFEE can be leveraged by any suitably adapted form compiler. For example, in Firedrake a modified version of the FEniCS Form Compiler producing abstract syntax trees instead of strings is used. COFFEE itself provides an interface for building such abstract syntax trees. In particular, COFFEE aims to decouple the form manipulation (i.e., the ``math level'') from code optimization (flops executed, low level optimization such as vectorization). Another viewpoint is: developers of form compilers should not worry about expression optimization in any way, since this is handled at a lower level.

@@

\subsection{Experimental Setup}

@@

\subsection{Performance Results}
\label{sec:perf-results-forms}

>>>>

\subsection{Quadrature mode}
Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies that a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Tensor contraction mode}
\label{sec:tc}
Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. Because of only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated ``reference tensor'' $A_{i_1 i_2 \alpha}$ and the cell-dependent ``geometry tensor'' $G_{K}^\alpha$ are exposed.

@@

\subsection{Loop nests and optimality}
In order to make the article self-contained, we start with reviewing basic compiler terminology.

@@

\subsection{Sharing elimination}
\label{sec:sharing-elimination}
We start with introducing the fundamental notion of sharing.

@@

\subsubsection{Normal form}
Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry depending on the differential operators employed. Note how the normal form respects Definition~\ref{def:fem-loopnest}. The following sections exploit the normal form to explain how we perform sharing elimination in finite element integration loop nests.

\subsubsection{Outer product loop nest}
\label{sec:se-opln}
In the special case in which test and trial functions are in the same function space, there exists $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$. That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$). Both $\alpha$ and $\beta$ can be hoisted at the level of the closest common loop ($L_i$) and the common sub-expressions be replaced with suitable temporaries. This has been detailed in~\citeN{Luporini}.

In the general case, instead, test and trial functions belong to different function spaces. By factorizing, in sequence, any test function $a_{ij} \in \alpha_{eij}$ and then any trial function $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops can be hoisted (a simplistic example is showed in Figure~\ref{code:multi_loopnest}).

These transformations minimize the operation count \textit{within} the outer product loop nest. However, when both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count. The difference in operation count tends however to be negligible, since only the outer loops are affected.

\subsubsection{Reduction and free order loop}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}

\subsubsection{Observations and underlying assumption}
\label{sec:se-assumptions}
What we are advocating is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that sub-expressions must be placed within the loops that they logically belong to (and not in a deeper level), and this can only be accomplished by systematic factorization and code motion.

There only are two pathological circumstances in which this approach might not lead to optimality: 
\begin{itemize}
\item As already explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes two alternative transformations.
\item The loops degenerate (i.e., their trip count is 1). This makes the factorizations described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} sub-optimal since code motion is now irrelevant. For example, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ and factorizing $L_i$-dependent symbols is optimal if and only if the size of $L_i$ is strictly greater than 1 (the sub-expression was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient).
\end{itemize}

\subsubsection{Algorithm}
Algorithm~\ref{algo:opt} describes how we manipulate the input loop nest and expression for performing sharing elimination. Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} can be seen as a constructive proof of optimality under the (reasonable) assumptions in Section~\ref{sec:se-assumptions}.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree.

We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice of what we should factorize first impacts the resulting operation count. It depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction.

Assume, without loss of generality, that we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.

For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted. Every time a sub-expression is hoisted, we search and replace all common sub-expressions. This last step is probably practically useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
\end{Algo}

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination reduces the operation count by applying three operators: factorization, code motion and common sub-expression elimination. A fourth operator impacting the operation count is \textit{reduction pre-evaluation}. In this section, we discuss its role and explain that there exists a single point in the whole transformation space (i.e., a specific factorization) that makes applying this operator legal.

@@

Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. However, we make use of a result -- the foundation of tensor contraction mode -- to simplify our task. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors. This result can be turned into a transformation algorithm for loops and expressions, for which we provide a succinct description below.

@@

\subsection{Transformation algorithm}

@@

\subsection{Automation through the COFFEE language}
COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality). In other words, COFFEE only exploits domain properties, and it is not specifically tied to finite element in any way. Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.

@@

\subsection{Independence from form compilers}
\label{sec:zeros}
One of the key pillars of COFFEE is independence from form compilers. Being capable of handling generic abstract syntax trees, COFFEE can be leveraged by any suitably adapted form compiler. For example, in Firedrake a modified version of the FEniCS Form Compiler producing abstract syntax trees instead of strings is used. COFFEE itself provides an interface for building such abstract syntax trees. In particular, COFFEE aims to decouple the form manipulation (i.e., the ``math level'') from code optimization (flops executed, low level optimization such as vectorization). Another viewpoint is: developers of form compilers should not worry about expression optimization in any way, since this is handled at a lower level.

@@

\subsection{Experimental setup}

@@

\subsection{Performance results}
\label{sec:perf-results-forms}
